---
title: "Lecture 6 in-class apply functions"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: inline
---



## Review 

Why not Excel??
    + no record of data manipulations  
    + hard to troubleshoot  
    + not possible for large datasets  

```{r eval=F}
#suppressPackageStartupMessages(library(tidyverse));
library(tidyverse)
library(data.table)
data_link <- "http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/BloodPressure.dat"
data_raw <- fread(data_link,sep = " ") 
################# without renaming
data_tidy1 <- data_raw[,-5] %>% 
    gather(key=device,value=reading,-Day) %>%
    extract(col=device,into="reader",regex="([[:alpha:]]+)",remove = F) %>% 
    extract(col=device,into="rep",regex="([[:digit:]]+)",remove = T)
################# with renaming
data_tidy2 <- data_raw[,-5] %>%
    setNames(c("Day",paste("Dev",1:3,sep="_"),paste("Doc",1:3,sep="_"))) %>%
    gather(key=device,value=reading,-Day) %>%
    separate(col = device,into = c("reader","replicate"),sep="_")
################# base R
data_tidy3 <- data_raw[,-5]
colnames(data_tidy3) <- c("Day",rep("D",6))
data_tidy3 <- rbind(data_tidy3[,c(1,2)],data_tidy3[,c(1,3)],data_tidy3[,c(1,4)],
                    data_tidy3[,c(1,5)],data_tidy3[,c(1,6)],data_tidy3[,c(1,7)])
colnames(data_tidy3) <- c("Day","reading")
data_tidy3$reader <- c(rep("Dev",15*3),rep("Doc",15*3))
data_tidy3$replicate <- rep(1:3,each=15)
```


## Problem 2 --

Create a function to find solutions to the function below using Newton's method.  The answer should include the solutions with tolerance used to terminate the loop and a plot showing the iterations on the path to the solution.

$$
f(x) = 2*sin(3*x)+(x-2)^2
$$

Should we plot it first to get an idea of where the roots should occure?
```{r}

```

OK, how does Newton's method work??  What we want are iteratations to:

$$
f(x_{i+1}) = x_i - \frac{f(x_i)}{f'(x_i)}  
$$

OK, so we need f'(x): 

```{r, eval=F}

update_X <- function(...){
    
}

need a start, stopping rule, iteration method, container for results
```

## Problem 3 --

Choosing $$x_0$$ was arbitrary, what if we choose a bunch of $x_0$'s?


```{r eval=F}


```